\section{Discussion}

The experimental framework established in this research serves to validate the 
feasibility of utilizing New York City’s existing traffic surveillance infrastructure
for high-resolution vehicle classification and emissions estimation. The results 
demonstrate that an optimized vehicle detection pipeline, specifically one centered
on the YOLOv11s architecture, can successfully overcome the visual challenges 
inherent in low-resolution, 352$\times$240 traffic imagery. This discussion 
synthesizes the technical performance of the evaluated models, the importance of 
the MOVES-based taxonomy, and the broader implications of these findings for 
scalable environmental health monitoring in dense urban environments.

% ==============================================================================

\subsection{Evaluation of Detection Architectures for NYC Imagery}

The selection of a primary detection model for this study was driven by a rigorous 
comparison across three distinct paradigms: one-stage convolutional neural networks,
two-stage networks, and transformer-based architectures. 
Here, one-stage detectors refer to models such as YOLO that predict object locations 
and classes directly from the image in a single pass, whereas two-stage models first 
propose candidate object regions and then classify them.
As the data in Table~\ref{tab:architecture-comparison-summary} 
indicates, the YOLO models performed consistently better than Faster R-CNN and the 
DETR models in terms of AP50 and mAP on the validation set.

\begin{table}[!htbp]
    \centering
    \caption{Baseline validation performance of the principal detector architectures evaluated in this study}
    \label{tab:architecture-comparison-summary}
    \begin{tabular}{lccc}
        \toprule
        \textbf{Model Architecture} & \textbf{AP50} & \textbf{mAP} & \textbf{Source} \\
        \midrule
        YOLOv11s         & 0.804 & 0.615 & Table~\ref{tab:yolo-baseline-results} \\
        YOLOv26s         & 0.802 & 0.632 & Table~\ref{tab:yolo-baseline-results} \\
        YOLOv8s          & 0.750 & 0.606 & Table~\ref{tab:yolo-baseline-results} \\
        Faster R-CNN     & 0.706 & 0.509 & Table~\ref{tab:faster-rcnn-results} \\
        RT-DETRv2        & 0.688 & 0.516 & Table~\ref{tab:detr-results} \\
        Deformable DETR  & 0.564 & 0.421 & Table~\ref{tab:detr-results} \\
        \bottomrule
    \end{tabular}
\end{table}

The performance difference in architectures suggests that when dealing with low-quality, 
352$\times$240 traffic camera images, having a highly efficient model design (like YOLO)
is actually more important than using a massive, complex architecture. Traditional 
two-stage detectors like Faster R-CNN try to locate candidate objects first and classify 
them second. However, across the 1,200-image dataset, this two-stage approach struggled 
to reliably detect the vehicles. Even when the Faster R-CNN model was tuned 
(the \verb|fasterrcnn_resnet50_fpn_v2| version) with 8 to 128-pixel anchors, it 
still couldn't beat the baseline performance scores (AP50) of the YOLO models.

Furthermore, the transformer-based models (the DETR family) demonstrated a clear 
sensitivity to data volume and augmentation strategy. As noted in Section~\ref{sec:detr},
Deformable DETR's performance improved significantly (from 0.428 to 0.564 AP50) when
the image augmentation pipeline was lightened. This suggests that these types of 
models struggle to learn from a small dataset if the training images are too heavily
modified or distorted. While newer variants like RT-DETRv2 showed promise, reaching 
an AP50 of 0.688, they still remained less accurate than the YOLOv11s model. 
Ultimately, YOLO's architecture proved more effective at extracting the necessary 
visual features from these low-resolution traffic scenes 
\cite{yolo11_ultralytics, ultralytics_rtdetrv2_yolo11}.

The performance advantage of YOLOv11s observed in this study likely stems from its 
ability to efficiently process objects at multiple scales. As outlined in 
Section~\ref{sec:yolov11}, its architecture combines visual features from different resolutions, 
which appears to help the model preserve semantic information for vehicles ranging 
from large buses in the foreground to small sedans in the distance. This multi-scale
capability is particularly critical for New York City’s traffic environment, where 
the high mounting of the cameras means a vehicle close to the lens appears massive, 
while the exact same type of vehicle further down the street appears as just a few 
pixels.

% ==============================================================================

\subsection{Rationale for the MOVES-Based Taxonomy}

The performance of the final detector was contingent not only on the architecture,
but on how the vehicle classes were defined. The MOVES-based taxonomy was created 
to balance what is actually visible in low-resolution traffic cameras with the data
required by the MOVES emissions software. Instead of forcing the model to distinguish
between vehicle body styles that look nearly identical at 352x240 pixels, this
approach grouped vehicles into broader categories based on their emissions profiles.

This strategy was applied from the very beginning of the data annotation process. 
Visually similar body styles like coupes and hatchbacks were labeled as \verb|sedan|,
while crossovers were labeled as \verb|suv|, relying on dependable visual cues 
like cabin height and rear profile sloping. In the final taxonomy, these labels were 
mapped directly to MOVES categories: sedans became \verb|passenger_car|, while SUVs,
pickup trucks, and minivans were grouped into \verb|passenger_truck|. This specific 
separation is critical for downstream environmental analysis, because EPA's MOVES 
framework treats passenger cars, passenger trucks, and light commercial trucks as 
distinct on-road source types, and computes criteria-pollutant emissions such as $NO_{x}$
and $PM_{2.5}$ through source-type-specific emission-rate tables 
\cite{us_epa_moves3_2020, USEPA_overview_2025}.

The dataset statistics in Table~\ref{tab:class-counts} highlight the dominance of 
passenger car and passenger trucks, which make up over 77\% of all identified-vehicle annotations. 
Under these conditions, the most significant risk to model performance was not a failure 
to detect vehicles, but a failure to distinguish between the two passenger vehicle 
categories. This visual overlap motivated the handling of the \verb|unknown_vehicle|
class, which was the single largest category in the dataset with 9,625 instances.

Using \verb|unknown_vehicle| as an ignore region rather than a normal training class
helped avoid forcing the model to learn from ambiguous examples. In supervised vision
tasks, noisy or uncertain labels can hurt generalization, and object-detection 
benchmarks commonly exclude regions that cannot be annotated reliably from scoring 
rather than treating them as standard objects \cite{song2022learningnoisylabelsdeep}. 
During training, these ambiguous regions were masked with a neutral fill value so
the detector would not learn from pixels that human annotators could not label 
confidently. As a result, the reported AP50 and mAP better reflected performance
on clearly identifiable vehicles instead of penalizing the model for inherently
uncertain regions.

% ==============================================================================

\subsection{Data-Centric Refinement vs. Model Complexity}

The improvement from the baseline YOLOv11s model (AP50 of 0.804) to the final selected 
version (AP50 of 0.857) is consistent with recent work emphasizing that model performance
can often be improved substantially through better data curation and refinement, even 
without changing the underlying architecture \cite{feuer2024selectlargescalebenchmarkdata, goyal2024scalinglawsdatafiltering}. 
The experiments in Section~\ref{sec:data-refinement}
suggest that the main limitation was not model capacity alone, but a concentrated 
weakness between predicting \verb|passenger_car| and \verb|passenger_truck|.

Manual inspection of the Figure~\ref{fig:val-cm-pre} and Figure~\ref{fig:val-cm-post}
confusion matrices revealed that while the model had learned to localize vehicles 
with high precision, it frequently swapped labels between sedans and SUVs.
The hard negative mining strategy in Section~\ref{sec:data-refinement} addressed this by identifying
100 images where these swaps occurred and duplicating them back into the training set.
This targeted intervention proved significantly more effective than broad hyperparameter
variation or general oversampling of rare classes like motorcycles.

Adding 400 extra images to the training set led to lower performance (AP50 of 0.776) 
than the targeted 100-image hard negative configuration (AP50 of 0.857). This shows 
that simply increasing dataset size is not enough. New data is most useful when it 
addresses the model’s existing failures. Here, the smaller hard negative set performed
better because it directly reinforced the difficult boundary between similar vehicle
classes. For urban monitoring applications, this suggests that iterative human-in-the-loop
data refinement is more effective than untargeted dataset expansion. By reviewing the 
model's own errors, researchers can improve the training set precisely where the detector 
is weakest.

% ==============================================================================

\subsection{Evaluation and Inference Results}

A key result from the threshold sensitivity analysis in Section~\ref{sec:conf-nms-analysis}
is that the validation metrics were driven much more by the confidence threshold 
than by the NMS IoU setting. Across all 36 settings in the sweep, the best performance
occurred at the lowest confidence threshold tested, \verb|conf| = 0.001 and 
\verb|det_iou| = 0.7, where AP50 reached 0.857 and mAP reached 0.661. As the 
confidence threshold increased, all three metrics fell steadily. At \verb|conf| = 0.25,
AP50 was still about 0.809--0.811 and mAP was about 0.622--0.626, which is only 
a moderate drop from the theoretical peak. In contrast, changing \verb|det_iou| 
from 0.4 to 0.7 had only a very small effect: the mean AP50 across all thresholds
changed from 0.780 to 0.781, while the mean mAP changed from 0.602 to 0.606. 
This shows that confidence filtering, not NMS tuning, was the dominant calibration 
variable.

This helps explain the large number of background false positives seen in the 
initial validation confusion matrices like Figure~\ref{fig:val-cm-pre}. 
Standard COCO-style AP evaluation rewards recall across the full precision-recall 
curve, so it naturally favors extremely low confidence thresholds such as 
\verb|conf| = 0.001 \cite{ultralytics_validation}. In practice, however, this setting
encourages the detector to keep many weak predictions, including boxes on ambiguous
background textures, which creates visual clutter and ghost vehicles. The heatmap 
results show that raising the threshold to \verb|conf| = 0.25 sacrifices only 
about 0.045 AP50 and about 0.035 mAP relative to the best setting, while greatly 
improving the practical usability of the detections. For downstream tasks such as 
ATR volume comparison, this makes confidence thresholding an essential deployment 
step, as maximizing AP at very low confidence is not automatically the setting 
that gives the cleanest and most reliable vehicle counts.

% ==============================================================================

\subsection{Generalization Stability and Domain Sensitivity}

A key question for practical deployment is how well the final detector generalizes 
beyond the data used for model selection. To examine this, the final YOLOv11s model 
was evaluated at two levels: first on the held-out test set drawn from the same 
12-camera pool as training and validation, and then on a separate cross-camera 
benchmark built from three entirely unseen cameras.

On the held-out test set, the model achieved an overall AP50 of 0.772 and an overall
mAP of 0.607. These results remained reasonably close to the validation performance, 
although both were lower than the validation values of AP50 of 0.857 and mAP of 0.661.
The drop of 0.085 in AP50 and 0.054 in mAP suggests that the validation improvements
were real, but that performance still decreased when the model was applied to a fully
separate split. In other words, the detector generalized well within the original 
dataset, but not without some loss.

The larger drop appeared on the cross-camera benchmark, where overall AP50 fell to 
0.658 and overall mAP fell to 0.488 (Table~\ref{tab:cross-camera-results}). 
Relative to the held-out test set, this corresponds to an additional decrease of
0.114 in AP50 and 0.119 in mAP. This result shows that the model transferred less 
effectively to unseen camera views than to new images from the original camera set
\cite{hsu2019progressivedomainadaptationobject, vis_domain_adapt_paper}. 
That benchmark also appeared harder visually than the training set. The unseen 
cameras showed broader urban scenes, more background content, and many vehicles 
that appeared smaller because they were farther from the camera. Broadway and 
Houston seemed especially difficult, since it captured a wider street view with 
more surrounding buildings and more distant vehicles.

\begin{table}[!htbp]
    \centering
    \caption{Comparison of held-out test-set and cross-camera benchmark performance for the final YOLOv11s model. The change column reports the difference between the cross-camera benchmark and the held-out test set ($\Delta = \text{cross-camera} - \text{held-out test}$).}
    \label{tab:cross-test-comparison}
    \resizebox{\textwidth}{!}{%
    \begin{tabular}{l c c c}
        \toprule
        Metric &
        \shortstack{Test Set} &
        \shortstack{Cross-Camera Benchmark} &
        \shortstack{Change ($\Delta$)} \\
        \midrule
        Overall AP50                         & 0.772 & 0.658 & -0.114 \\
        Overall mAP                          & 0.607 & 0.488 & -0.119 \\
        Passenger Car AP50                   & 0.779 & 0.626 & -0.153 \\
        Passenger Truck AP50                 & 0.868 & 0.761 & -0.107 \\
        Light Commercial Truck AP50          & 0.867 & 0.685 & -0.182 \\
        Truck AP50                           & 0.926 & 0.735 & -0.191 \\
        Bus AP50                             & 0.895 & 0.635 & -0.260 \\
        Motorcycle AP50                      & 0.298 & 0.505 & \phantom{-}0.207 \\
        \bottomrule
    \end{tabular}%
    }
\end{table}
\FloatBarrier

The Table~\ref{tab:cross-test-comparison}. class-level AP50 differences from test
set to the cross-camera benchmark show that the performance decline was not uniform. 
Passenger trucks were the most stable major class, while passenger cars, light 
commercial trucks, trucks, and especially buses were more sensitive to the unseen 
views. This pattern suggests that cross-camera generalization depended not just 
on object size, but on how consistently each class appeared across camera perspectives.
Passenger trucks seem to have retained enough distinctive visual structure to 
transfer relatively well, whereas buses and light-duty classes were more disrupted 
by the new viewpoints and scene composition. The bus decline was especially notable,
indicating that strong performance on the original camera set did not automatically 
carry over to substantially different views. Motorcycle AP50 increased on the 
cross-camera benchmark, but that result should be interpreted cautiously because 
it was based on only two labeled motorcycles, making it too small a sample to 
support a broader conclusion.

Taken together, these results show two levels of generalization. The model remained 
strong on the held-out test set, which indicates that the selected improvements were 
not limited to the validation split alone. However, performance dropped more noticeably
on the unseen-camera benchmark, showing that transfer to new camera views was still 
a harder problem. For deployment, this suggests that the detector can still be used 
out of the box for broad traffic monitoring, but more accurate class-specific 
estimation would likely require adding labeled example images from each new location 
to the training set. In particular, brief site-specific fine-tuning would likely 
help the detector adapt to new scene layouts, viewing distances, and vehicle appearance 
at that camera.

% ==============================================================================

\subsection{Analysis of Traffic Count Agreement}

The Section~\ref{sec:atr-analysis} ATR comparison was used to test whether the 
detection pipeline could recover not only the daily pattern of traffic, but also 
the overall scale of vehicle activity. To evaluate this, three metrics were used: 
Pearson correlation, mean absolute error (MAE), and mean absolute percentage error 
(MAPE). In the context of this study, Pearson correlation measures how well the 
detector matched the shape of the hourly traffic pattern, regardless of total 
count magnitude. MAE measures the average hourly difference in raw vehicle counts. 
MAPE measures the average hourly error as a percentage of the ATR count, which 
makes it easier to compare error across locations with different traffic volumes.

As observed from Table~\ref{tab:atr-count-comparison}, the closest agreement was 
at Atlantic Avenue at Boerum Place. On March 18, the detector produced a daily 
total of 7,559 vehicles compared with an ATR total of 8,725, with a Pearson 
correlation of 0.844, an MAE of 82.67 vehicles per hour, and a MAPE of 22.96\%. 
On March 19, agreement remained fairly strong, with 7,060 YOLO-derived vehicles 
versus 9,188 ATR vehicles, a Pearson correlation of 0.803, an MAE of 104.17, and 
a MAPE of 27.28\%. These were the best results in the comparison, suggesting that
at this location the detector captured both the hourly trend and the total volume
reasonably well.

The other two locations showed a different pattern. At Pelham Parkway Westbound 
at Boston Road, Pearson correlation was consistently strong, ranging from 0.780 to
0.906, which indicates that the detector generally tracked the timing of higher-
and lower-volume hours well. However MAE ranged from 421.04 to 644.83 vehicles 
per hour, and MAPE ranged from 70.66\% to 78.47\%. This means that the model
captured the daily traffic shape but substantially undercounted the true number 
of vehicles. Broadway at Houston showed the weakest and least consistent agreement 
overall. Pearson correlation ranged from 0.234 to 0.753, while MAPE remained high 
on all four days, from 67.22\% to 69.08\%. The worst single-day result occurred 
on March 16, when the detector counted 3,431 vehicles compared with 12,382 from
the ATR, with a Pearson correlation of only 0.234. This indicates that at Broadway 
and Houston, the method sometimes failed to recover even the basic hourly pattern 
reliably.

These results show a clear difference between pattern agreement and count agreement.
High Pearson correlation, as seen especially at Pelham Parkway, means the detector 
was often able to identify when traffic was relatively busy or quiet over the course 
of the day. High MAPE and MAE, however, show that the absolute count scale could 
still be far off. For this project, that distinction matters. If the goal is to 
estimate temporal traffic dynamics, such as peak hours or relative changes in flow,
then strong correlation is useful. But if the goal is to estimate total vehicle 
activity for emissions modeling, then large magnitude errors remain a serious limitation.

Several factors likely contributed to this undercounting. First, the camera-based 
method only sampled one image every two minutes, so vehicles passing between snapshots 
were never observed. Second, the detector could only count vehicles that were visible 
within the camera's field of view, unlike ATR sensors that record traffic continuously
across the roadway. Occlusion likely also played a role, especially when larger vehicles
blocked smaller ones. More generally, this comparison was necessarily approximate 
because the ATR counts came from 2025 while the camera images came from 2026. The 
matched dates were therefore the best available comparison, but they were not exact
same-day ground truth. In addition, several of the matched days fell during the week
of Saint Patrick's Day, so location-specific events, celebrations, or traffic 
disruptions may have caused year-to-year differences that were unrelated to detector 
performance.

Overall, the ATR comparison suggests that the pipeline was more successful at recovering 
the relative hourly structure of traffic than at recovering exact vehicle totals. 
Atlantic Avenue showed that closer agreement was possible in a favorable view, while 
Pelham Parkway showed that strong temporal tracking could coexist with severe 
undercounting. Broadway and Houston demonstrated that in more difficult scenes,
even the hourly pattern could break down. For practical use, this means the detector
can provide useful traffic-pattern information, but absolute count estimates should
be interpreted cautiously unless additional calibration is performed.

% ==============================================================================

\subsection{Implications for Public Health and Environmental Policy}

The results of this study suggest that existing traffic-camera networks could 
support a more spatially detailed approach to traffic-related air-quality analysis.
In New York City, traffic contributes about 14\% of citywide $PM_{2.5}$, and prior
work has shown that trucks and buses account for a particularly large share of 
traffic-attributable ambient $PM_{2.5}$ in the city. Within that context, the main
contribution of this pipeline is not direct pollution measurement, but the ability
to estimate vehicle composition at specific intersections using a MOVES-aligned taxonomy.
That is potentially useful because fixed air-quality monitors provide only limited 
spatial coverage, whereas camera-based classification could help identify locations 
with persistently high proportions of buses, trucks, or other higher-emitting vehicle
groups that are especially relevant for street-level exposure.

At the same time, the results also show that this approach should be treated as a 
screening and prioritization tool rather than a full replacement for conventional 
monitoring. The detector performed strongly on the held-out test set, but cross-camera 
performance declined on unseen views, and the ATR comparison showed that hourly 
traffic patterns were recovered more reliably than exact count magnitude. In practical
terms, this means the pipeline is better suited to highlighting relative differences
across locations, time periods, or vehicle mixes than to producing perfectly calibrated
traffic totals without local adjustment. Even so, that kind of intersection-level 
screening could still be useful for public-health and planning decisions, such as
identifying corridors where freight or bus activity is concentrated and where 
follow-up interventions or measurements may be warranted. More broadly, the benefit
of this approach is that it can build on infrastructure already being expanded by 
NYC DOT, which has described plans to scale connected-vehicle and smart-intersection
deployments from about 150 to as many as 600 intersections. If paired with limited 
site-specific calibration, the pipeline could offer a relatively low-cost and 
privacy-preserving way to add transportation exposure intelligence to existing 
urban traffic systems.

% ==============================================================================

