\section{Experimentation and Results}

Vehicle detection experiments in this study were centered on a 6-class taxonomy 
designed to support emissions-oriented analysis. Several preliminary configurations 
were explored during early development, but those runs were primarily used to 
understand the difficulty of the task and refine the problem formulation. Thus,
the results presented in this section focus on experiments conducted under the 
final 6-class setting. 

First, YOLO-based experiments are presented, including comparisons across model
variants and refinements through data-centric strategies such as hard negative 
mining and dataset refinement. Other detectors like DETR-based models and Faster
R-CNN were examined to compare architectures within the project’s context. 

After identifying the strongest detector, additional experiments evaluate its 
robustness through threshold and NMS sensitivity analysis, cross-camera generalization,
and agreement between predicted vehicle counts and reference ATR or manual counts. 
Together, these experiments exhibit the practical suitability of the final system 
for traffic-monitoring applications.

% ==============================================================================

\subsection{MOVES-based Taxonomy}

For the experiments reported in this chapter, model training and evaluation were 
conducted using a final 6-class taxonomy: 
\verb|passenger_car|, \verb|passenger_truck|, \verb|light_commerical_vehicle|, 
\verb|bus|, \verb|truck|, and \verb|motorcycle|. 
This label set was selected to provide a practical balance between compatibility
with downstream MOVES-based emissions analysis and what could be distinguished 
reliably from low-resolution traffic-camera imagery. Rather than attempting to 
reproduce the full MOVES source type framework directly at the detector level, 
the taxonomy in this study was designed as a MOVES-aligned grouping that preserved 
broad vehicle categories relevant to emissions modeling while reducing distinctions
that were difficult to learn consistently from the available images.
